---
title: "Workshop_demo"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installation

```{r echo = FALSE, message = FALSE, warning = FALSE}
# install.packages("dplyr")
library(dplyr)
```

## Data set

``` {r echo = FALSE}
starwars 
head(starwars)
dat <- starwars
``` 

## Pipe operator 

```{r echo = FALSE}
# Ugly example 

dat1 <- select(dat, c(name, height, mass))
dat1_1 <- filter(dat1, height >= 100)
View(dat1_1)

# Beautiful example 

dat1 <- dat %>% select(name, height, mass) %>% 
                filter(height >= 100) %>% 
                View()
```

## filter() 

Purpose: select a subset of rows in a data frame 
Structure: filter(tibble/data frame, variable of choice and expression)

``` {r}
# Filter for numerical values and character values
filter(dat, height >= 100)
dat %>% filter(height >= 100, 
               eye_color == "yellow") %>% 
        View()

# Filter for A OR B
dat %>% filter(height >= 100 | eye_color == "yellow") %>% 
        View()

# Filter for several values of the same type 
dat %>% filter(hair_color %in% c("blond", "brown")) %>% 
        View()

# Filter out cases
dat %>% filter(!(height >=100)) %>% 
        View()
``` 

## slice 

Purpose: index rows by their (integer) locations 
Structure: slice(tibble/data frame, location)

```{r }
# First 5 rows
slice(dat, 1:5)

# slice_head() for rows at the top of the df 
# n refers to amount of desired observations
slice_head(dat, n = 3)

# slice_tail() for rows at the bottom of the df 
dat %>% slice_tail(n = 4)

# slice_sample() for random selection 
# prop for proportion of cases
dat %>% slice_sample(n = 4) 
dat %>% slice_sample(prop = 0.1)

# slice_max() for rows with highest values of a variable 
dat %>% slice_max(height, n = 3)

# slice_min() for rows with lowest values of a variable 
dat %>% slice_min(mass, n = 3)
```


## arrange()

Purpose: reorders rows 
Structure: arrange(df/tibble, variable)

``` {r}
# Default is ascending
dat %>% slice(1:5) %>% 
        arrange(mass) %>% 
        View()

# Descending 
dat %>% slice(1:5) %>% 
        arrange(desc(mass)) %>% 
        View()
``` 



## select()

Purpose: create a subset of a df by selection of columns 
Structure: select(dt/tibble, columns of interest)

```{r echo=FALSE}
dat1 <- slice(dat, 1:5)

# Select by name of columns 
dat1 %>% select(name, hair_color) %>% 
        View()

dat1 %>% select(name:hair_color) %>% 
        View()

# Exclude by name 
dat1 %>% select(-mass, -hair_color) %>% 
        View()
dat1 %>% select(!(mass:hair_color)) %>% 
        View()

# Select columns by feature 
dat1 %>% select(contains("color")) %>% 
        View()
dat1 %>% select(-ends_with("color")) %>% 
        # opposite is starts_with()
        View()

# Rename variables 
# Not recommended because it drops all other variables 
dat1 %>% select(fabulous_name = name) %>% 
        View()

# Reorder 
dat1 %>% select(name, hair_color, skin_color, everything()) %>% 
        View()
```

## rename() 
Purpose: rename columns
Structure: rename(df/tibble, new_colname = colname)

```{r}
dat1 %>% rename(fabulous_name = name) %>% 
        View()
```


## mutate() 
Purpose: add new columns that are functions of existing columns 
Structure: mutate(new_colname = function)

``` {r}
# Constructing a BMI column
dat1$height <- dat1$height/100
dat1 %>% mutate(BMI = mass/(height^2)) %>% 
        select(name, BMI, everything()) 

# transmute() to only keep the new variables
# Instead of separately preparing the height measure, you can
# also include it in the function 
dat1 %>% transmute(
        height_m = height / 100, 
        BMI = mass / (height_m^2)
)

# Unnamed columns 
dat %>% slice(1:5) %>% 
        mutate(height + 10) 
``` 


## summarise() 
Purpose: collapse df to single variables
Structure: summarise/summarize(df, varname = function)

```{r}
# Only getting mean height 
dat1  %>% summarise(
        mean_height = mean(height, na.rm = TRUE), 
        mean_mass = mean(mass, na.rm = TRUE)
        )

# Variation: summarise_at() to affect variables selected with a 
# character vector or vars()
dat %>% summarise_at(vars(c(2, 7)), mean, na.rm = TRUE)
dat %>% summarise_at("mass", mean, na.rm = TRUE)
dat %>% summarise_at(vars(height:mass), mean, na.rm = TRUE)

# Variation: summarise_if() affects variables selected with a 
# predicate function 
dat %>% summarise_if(is.numeric, mean, na.rm = TRUE)
```

## group_by()
Purpose: create groups 
Structure: group_by(df, variable)

``` {r}
# Instead of the mean height across all creatues, let's focus mean height
# for each species 
dat %>% group_by(species) %>% 
        select(height, mass) %>% 
        summarize(
                mean_height = mean(height, na.rm = TRUE), 
                mean_mass = mean(mass, na.rm = TRUE)
        )
``` 

## count() 
Purpose: count the unique values of one or more variables
Structure: count(df, var) 

```{r}
# Counting the rows by species 
dat %>% count(species)

# Alternative way: n() which returns group size 
dat %>% group_by(species) %>% 
        summarize(n = n())

# Count the frequency of weights (mass) for each gender
dat %>% count(gender, wt = mass)

```

## EXTRA: grep(), gsub(), gather(), spread() 

``` {r}
## grep() 
?grep
## Structure: grep(pattern, vector)
grep("blond", dat$hair_color) #length()
dat[grep("blond", dat$hair_color),]
dat[,grep("color", colnames(dat))]


## gsub() 
?gsub
## Structure: gsub(pattern_to_be_replaced, new_pattern, df)
gsub("brown", "brunette", dat$hair_color)

dat_na <- dat[!(is.na(dat$hair_color)|dat$hair_color == ""),]
dat_na %>% mutate_at(vars(hair_color), ~gsub("brown", "brunette",.)) %>% 
        View()
# What is the tilde sign doing here? Dplyr reserves the use of it 
# for non-standard evaluation (NSE), i.e. NS functions 

library(tidyr)

## pivot_longer(), formally known as gather()
?pivot_longer
dat_long <- dat %>% pivot_longer(cols = ends_with("color"),
                        names_to = "col", "value") %>% 
                        select(name, col, value, everything()) 
View(dat_long)


## pivot_wider(), formally known as spread()
?pivot_wider
dat_long %>% pivot_wider(names_from = col, 
                    values_from = value) %>% 
                    select(name, hair_color, eye_color, 
                    skin_color, everything()) %>% 
                    View()
``` 